---
title       : Practical Data Wrangling
subtitle    : DataDive 101 // 2014-11-15
author      : Eugene Teo
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : solarized_light     # tomorrow
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

<q>Residents in Jurong West saw a strange object in the sky on Saturday night, describing it an unidentified flying object that hovered for as long as 15 minutes over some blocks of flats.</q>

---

## Data wrangling in 20 minutes!

We will walk through an interesting example from the book [Machine Learning for Hackers](http://shop.oreilly.com/product/0636920018483.do). We will look at the documented reports of UFO sightings in China.

You will need to clone my [datadive-101](https://github.com/eugeneteo/datadive-101) git repository so that you can follow along.

```{bash}
$ git clone https://github.com/eugeneteo/datadive-101.git
Cloning into 'datadive-101'...
remote: Counting objects: 58, done.
remote: Compressing objects: 100% (46/46), done.
remote: Total 58 (delta 13), reused 44 (delta 2)
Unpacking objects: 100% (58/58), done.
Checking connectivity... done.
```

```{bash}
$ shasum datadive-101/ufo_awesome.tsv.tar.bz2 
b1f58dfc72b3ba22286518efd0f0e0aac8f5a755  datadive-101/ufo_awesome.tsv.tar.bz2
```

---

## Install the necessary R packages

We will use `R.utils` to unarchive the dataset, and use `dplyr` to manipulate data frame-like objects.

```{r}
pkg <- c("dplyr", "R.utils")
new.pkg <- pkg[! (pkg %in% installed.packages())]
if (length(new.pkg)) {
  install.packages(new.pkg)
}

library(R.utils)
library(dplyr)
```

---

## Unarchive the dataset

```{r, echo=TRUE}
tardataset <- "./ufo_awesome.tsv.tar"
bzdataset <- gsub("$", ".bz2", tardataset)
dataset <- gsub(".tar$", "", tardataset)
if (! file.exists(dataset)) {
  if (! file.exists(bzdataset)) {
     # ufo_awesome.tsv's shasum: 39cfc99c9814d8806526a1adeaf3c582e47b0f27
     file.url <- "https://raw.githubusercontent.com/eugeneteo/datadive-101/master/ufo_awesome.tsv.tar.bz2"
     download.file(file.url, destfile = bzdataset, method = "curl")
  }
  bunzip2(bzdataset)
  untar(tardataset)
  file.remove(tardataset)
}
```

---

## Learn more about the dataset

```{bash}
$ file ufo_awesome.tsv
ufo_awesome.tsv: ASCII English text, with very long lines
```
```{bash}
$ wc -l ufo_awesome.tsv
61393 ufo_awesome.tsv
```
```{bash}
$ head -1 ufo_awesome.tsv | recode html..ascii
19951009  19951009	 Iowa City, IA			Man repts. witnessing "flash, followed
  by a classic UFO, w/ a tailfin at back." Red color on top half of tailfin.
  Became triangular.
```
```{bash}
$ head -1 ufo_awesome.tsv | od -c | grep '\\t'
0000000    1   9   9   5   1   0   0   9  \t   1   9   9   5   1   0   0
0000020    9  \t       I   o   w   a       C   i   t   y   ,       I   A
0000040   \t  \t  \t   M   a   n       r   e   p   t   s   .       w   i
```

---

## Read the .tsv file

```{r, echo=TRUE}
ufo <- tbl_df(read.delim(dataset, sep = "\t", header = FALSE,
                         stringsAsFactors = FALSE, na.strings = ""))
names(ufo) <- c("DateOccurred", "DateReported", "Location", "ShortDescription",
                "Duration", "LongDescription")
head(ufo)
```

---

## Format the dates

Convert the dates from `%Y%m%d` to `%Y-%m-%d` format.

```{r, eval=FALSE,echo=TRUE,results='hide',warning=FALSE,message=FALSE,error=FALSE}
ufo <- mutate(mutate(ufo, DateOccurred = as.Date(DateOccurred, format = "%Y%m%d")),
              DateReported = as.Date(DateReported, format = "%Y%m%d"))
```

A better way of writing this is to use the `%>%` operator. It turns `x %>% f(y)` into `f(x, y)`.

```{r, echo=TRUE}
ufo <- ufo %>%
  mutate(DateOccurred = as.Date(DateOccurred, format = "%Y%m%d"),
         DateReported = as.Date(DateReported, format = "%Y%m%d"))
```

Why is there an error?

---

## Not all dates are eight characters long

```{r, echo=TRUE}
head(ufo[ # row
  which(nchar(ufo$DateOccurred) != 8 | 
          nchar(ufo$DateReported) != 8)
  , 1]) # column
```

---

## Let's fix the dates

```{r, echo=TRUE}
good.rows <- ifelse(
  nchar(ufo$DateOccurred) != 8 | nchar(ufo$DateReported) != 8,
  FALSE, TRUE # returns FALSE if != 8
  )
head(good.rows)

table(good.rows)

ufo <- ufo[good.rows, ]
```

---

## Format the dates, take two

Replace the two existing variables (columns) with the correct date formatting.

```{r, echo=TRUE}
ufo <- ufo %>%
  mutate(DateOccurred = as.Date(DateOccurred, format = "%Y%m%d"),
         DateReported = as.Date(DateReported, format = "%Y%m%d"))
head(ufo)
```

---

## Show us only the UFO sightings in China

`gsub(pattern, replacement, x)`

`"^ "` matches the leading whitespace.

strsplit() will throw an error if the split character is not matched.

```{r, eval=FALSE,echo=TRUE,results='hide',warning=FALSE,message=FALSE,error=FALSE}
get.location <- function(loc) {
  loc <- gsub("^ ", "", loc)
  
  split.location <- tryCatch(
    strsplit(loc, ",")[[1]], error = function(e) return (c(NA, NA))
    ) # [1] "Nanjing (China)"
  
  split.location <- tryCatch(
    strsplit(split.location, " ")[[1]], error = function(e) return (c(NA, NA))
    ) # [1] "Nanjing" "(China)"
```

---

## Show us only the UFO sightings in China

`"^\\("` matches the first character `'('`.

`")$"` matches the last character `')'`.

```{r, eval=FALSE,echo=TRUE,results='hide',warning=FALSE,message=FALSE,error=FALSE}
  # [1] "Nanjing" "(China)"
  clean.location <- gsub("^\\(", "", split.location)
  clean.location <- gsub(")$", "", clean.location)
  # [1] "Nanjing" "China"

  if (length(clean.location) == 2) {
    return (clean.location)
  } else {
    return (c(NA, NA))
  }
}
```

---

## Apply get.location() on all locations

`lapply()` will apply the `get.location()` over the `ufo$Location` character vector.

```{r}
province <- lapply(ufo$Location, get.location)
head(province, 3)
```

---

## Convert the list into a matrix

Call `rbind()` on `province` list to convert it into a two-column matrix with the province data as the leading column.

```{r}
location.matrix <- do.call(rbind, province)
head(location.matrix, 8)
```

---

## Clean up and filter the dataset

Create two new variables for the province and country. To make the dataset more readable, remove the `Location` and `Duration` variables. Finally, filter the dataset to only display the documented reports from China. At the same time, make sure that the reports were not reported before they were occurred.

```{r, echo=TRUE}
new.ufo <- ufo %>%
  # Create two new columns for the province and country
  mutate(Province = location.matrix[, 1], Country = location.matrix[, 2],
         stringsAsFactors = FALSE) %>%
  select(-Location, -Duration) %>%
  # Exclude "South China, ME" as it is not in China
  filter(Country == "China" & Province != "South") %>%
  mutate(Mean = (DateReported - DateOccurred)) %>%
  filter(Mean >= 0)
```

---

## After cleaning up

```{r, echo=TRUE}
head(new.ufo, 10)
```

---

## Sightings in each province

Display the number of sightings in each province, and the average number of days the reporters took to report an incident.

```{r, echo=TRUE}
new.ufo %>%
  group_by(Province) %>%
  summarise(Sightings = n(), Avg.Reporting = mean(Mean)) %>%
  arrange(desc(Sightings)) %>%
  head(5)
```

---

## Sightings by provinces and short descriptions

```{r, echo=TRUE}
new.ufo %>%
  group_by(Province, ShortDescription) %>%
  summarise(Avg.Reporting = mean(Mean), Sightings = n()) %>%
  arrange(desc(Avg.Reporting)) %>%
  head(7)
```

---

## Look at China broadly

```{r, echo=TRUE}
new.ufo %>%
  group_by(ShortDescription) %>%
  summarise(Avg.Reporting = mean(Mean), Sightings = n()) %>%
  arrange(desc(Avg.Reporting)) %>%
  head(8)
```

---

## The second oldest sighting!

```{r}
oldest <- new.ufo %>%
  arrange(DateOccurred)
print(oldest$LongDescription[2])
```

```{bash}
$ cat /tmp/delme.txt | recode html..ascii
# [1] "the diameter looks 50m approximation.in the center,looks like black
#   clouds,and  something like rainbow around it.i was ready to go home with
#   others at that time.when i was getting up the bus,i heard somebody speak
#   alound"look!,look!".and i found they were looking up the sky.there
#   is the thing i beleive it is an UFO. it is beautiful and mystery.beause my
#   english is poor,i can not tell the thing very clear.but trust me,it is great
#   really."
```

---

## Other possible cleanups

```{r, echo=TRUE}
head(ufo$Duration, 20)
```

```{r, echo=TRUE}
tail(ufo$Duration, 20)
```

---

## References

1. Machine Learning for Hackers <http://shop.oreilly.com/product/0636920018483.do>
2. Example code from the book <https://github.com/johnmyleswhite/ML_for_Hackers>
3. Introducing dplyr <http://blog.rstudio.org/2014/01/17/introducing-dplyr/>
4. CRAN Task View: Web Technologies and Services <https://github.com/ropensci/mappingtools>

---

## Q&A

Any questions?
